Pipit: Reactive Systems in \fstar ~(Extended Abstract)

### 2023-05-10 2023pipit/Main.tex
pipit

Our contributions are as follows:
\begin{itemize}
  \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
  \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
    % TODO: CUT WAFFLE
    This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
    Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
  \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
\end{itemize}

@@

One common application in controllers is to implement filters, such as a \emph{finite impulse response} (FIR) filter.
An FIR filter takes an input signal and a vector of coefficients, and computes the dot product of the coefficients and the most recent values of the signal.
For discrete signals, it is equivalent to a weighted moving average.

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ fir (coefficients: list $\RR$) (s: stream $\RR$): stream $\RR$ = \\
    \> @match@ coefficients @with@ \\
    \> @|@ $[]$ \> $\to 0$ \\
    \> @|@ $c :: \cs $ \> $\to (s \cdot c) + \mbox{fir~} \cs ~(\xfby{0}{s})$ \\
\end{tabbing}

@@

\TODO{Required for submission: writing}
\begin{itemize}
  \item introduction
  \item running example: FIR?
  \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
\end{itemize}
\TODO{Required for submission: implementation}
\begin{itemize}
  \item finish proof of system correctness
  \item state end-to-end / capstone theorem that k-inductively verified systems apply to original semantics
  \item implement unverified CSE pass
  \item verify running example, with CSE and k-induction
\end{itemize}

Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
The original Kind \cite{hagen2008scaling}.
\CITEME{2022 thesis on Agda semantics of Lustre?}

Past-time modal temporal logic \cite{lichtenstein1985glory}
Lustre support for temporal logic \cite{halbwachs1993executable}

>>>>

In this extended abstract, we will briefly describe the following preliminary results, which we intend to publish pending further evaluation:
\begin{itemize}
  \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
  \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
    % TODO: CUT WAFFLE
    This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
    Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
  \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
\end{itemize}

@@

A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
An FIR filter takes an input signal and a vector of coefficients, and computes the dot product of the coefficients and the most recent values of the signal.
We can implement an FIR filter in Pipit as follows:

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ fir ($\textit{coefficients}$: list $\RR$) ($s$: stream $\RR$): stream $\RR$ = \\
    \> @match@ $\textit{coefficients}$ @with@ \\
    \> @|@ $[]$ \> $\to 0$ \\
    \> @|@ $c :: \cs $ \> $\to (s \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~s)})$
\end{tabbing}

Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals here, while the signal $s$ is a \emph{stream} of reals, and the result is the filtered stream.
The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($s \cdot c$) and delay the result ($\xfby{0}{\ldots}$) of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ s$), summing the two parts.
When there is no previous value, as is the case at the start of execution, the delay is given the initial value of zero.

Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as different tools support different features.
To write this program in Lustre, we would likely need to reformulate the program to use arrays: Vélus does not support arrays, so verified compilation is out; Kind2 uses a custom syntax for arrays that no compiler supports, so model-checking with Kind2 is out; and Lustre v6 does support arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous}, only supports reasoning with boolean abstractions.

\newcommand\ii{\textit{input}}

Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}).
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program, which can be easily executed in statically-bounded runtime and space.
For example, the weighted moving average $\mbox{fir}~[0.5; 0.3; 0.2] \ii$ becomes $(0.5 \cdot \ii) + \xfby{0}{((0.3 \cdot \ii) + \xfby{0}{((0.2 \cdot \ii) + \xfby{0}{0})})}$.

@@

% \TODO{Required for submission: writing}
% \begin{itemize}
%   \item introduction
%   \item running example: FIR?
%   \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
% \end{itemize}
% \TODO{Required for submission: implementation}
% \begin{itemize}
%   \item finish proof of system correctness
%   \item state end-to-end / capstone theorem that k-inductively verified systems apply to original semantics
%   \item implement unverified CSE pass
%   \item verify running example, with CSE and k-induction
% \end{itemize}

% Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
Cite: the original Kind \cite{hagen2008scaling}.
\CITEME{2022 thesis on Agda semantics of Lustre?}
Past-time modal temporal logic \cite{lichtenstein1985glory}
Lustre support for temporal logic \cite{halbwachs1993executable}
